Distinguishing among coalescent models using two-site allele frequency spectra

### 2018-08-16 manuscript/manuscript.tex
Line edits to manuscript.

One approach to identifying multiple mergers in genomic data is to use the SFS as a summary statistic.
\cite{BirknerEtAl2013, BlathEtAl2016, SpenceEtAl2016} derived methods for computing the expected site frequency spectrum of (simultaneous-)multiple merger coalescents.
\cite{EldonEtAl2015} showed that it is possible to distinguish between a multiple merger coalescent of the beta family and the Kingman coalescent with exponential growth using the SFS.
\cite{RodelspergerEtAl2014} detected non-Kingman signatures of widespread linked selection in the nematode \textit{Pristionchus pacificus} by demonstrating that the site frequency spectrum is non-monotonic, a signature of multiple mergers \autocite{NeherHallatscheck2013, BirknerEtAl2013}.

@@

Following the notation of \cite{Fu1995}, the site frequency spectrum of a sample of $n$ haploid genomes is $\left\{ \xi_i : 1 \leq i < n \right\}$, where $\xi_i$ is the fraction of sites containing a mutation with derived allele count $i$ in the sample.
In many cases, the ancestral allele is unknown and so the allele in $i$ samples and the complimentary allele in $n-i$ samples are indistinguishable.
Therefore, we will mostly consider the \textit{folded} site frequency spectrum $\left\{ \eta_i = \xi_i + (1-\delta_{i,n-i}) \xi_{n-i}: 1 \leq i \leq \floor{n/2} \right\}$, where $\delta_{k,k'}$ is the Kronecker delta. %, where $\eta_i = \xi_i + \delta_{n-2i} \xi_{n-i}$.
The SFS and folded SFS can be calculated from a set of single nucleotide polymorphisms (SNPs) without any information about their relative locations in the genome.

@@

\cite{BirknerEtAl2013} extended Fu's calculation to a family of multiple merger coalescents called beta coalescents.
This one-parameter family interpolates between the Kingman coalescent and the Bolthausen-Sznitman coalescent as the parameter, $\alpha$, varies from 2 to 1.
Beta coalescents arise in models with fat-tailed offspring distributions \autocite{Schweinsberg2003}, and the Bolthausen-Sznitman coalescent is the limiting distribution of genealogies in rapidly adapting populations \autocite{NeherHallatscheck2013}.
Like Fu, Birkner et al.\ were primarily concerned with computing the sample variance of SFS-based summary statistics such as Tajima's D \autocite{Tajima1989}.
As a result, they were mostly interested in the diagonal terms of the SFS covariance matrix, which dominate that calculation.
However, Figures 5 and 6 of \cite{BirknerEtAl2013} show positive correlations between $\xi_i$ and $\xi_j$ for $j \neq i, n-i$.
Thus, unlike the standard Kingman coalescent, the beta coalescent can generate positive associations between mutations with different minor allele counts.

@@

In our setting, the 2-SFS may be interpreted as a joint probability mass function over allele frequencies.
Given a coalescent model, the minor allele count at an arbitrary site is a random variable with probability mass function $p(i) = \E{\eta_i}$, where we define $\eta_0$ to be the fraction of monomorphic sites.
Similarly, the minor allele counts at a pair of sites separated by $d$ bases is a pair of random variables with joint probability mass function $p_d(i,j) = \E{\eta_{ij}(d)}$.
Comparing, \eq{eq:fpmi} with \eq{eq:pmi} shows that $\fpmi$ is the standard pointwise mutual information for pairs of allele frequencies.

This transformation of the 2-SFS has several useful properties.
First, because it is based on the minor allele frequencies, we may compute $\fpmi$ from data without knowing the ancestral allele.
Moreover, the results of \cite{Fu1995} show that $\fpmi(i,j) < 0$ for $i\neq j$ for non-recombining sites under the time-homogeneous Kingman coalescent.

@@

\subsection*{Weighted $\fpmi$}
For plotting purposes, we will also use a weighted version of the $\fpmi$:
\begin{align}
    \wfpmi(i,j:d) = \frac{\E{\eta_{ij}}}{\mu^2 \E{T_2}^2} \fpmi(i,j;d),
    \label{eq:wfpmi}
\end{align}
where $T_2$ is the coalescence time for a sample of size two.
The numerator of the weighting factor in \eq{eq:wfpmi} serves to emphasize the most common pairs of minor allele counts.
The denominator, which is proportional to the expected pairwise diversity, $\Pi$, ensures that $\wfpmi$ is invariant under changes in the overall mutation and coalescence rates.

@@

For each coalescent model, we simulated at least $10^4$ independent infinite-sites loci, each with total recombination rate, $r$ (See SI for parameter combinations).
We chose values of $r$ so that the combined parameter $r\E{T_2}$ varied over several orders of magnitude, where $\E{T_2}$ was measured from the simulation output.
For each locus, we measured $\{\tau_i : i = 1,\ldots,n-1\}$ in the two genealogies at each end of the loci.
We then calculated the expectations $\{\E{\tau_i\}}$ and $\{\E{\tau_{ij}}\}$ by averaging over independent loci.
These expectations allow us to calculate all of the 2-SFS statistics defined above.

@@

We simulated a model of recurring selective sweeps using the software \texttt{SLiM} \autocite{MesserEtAl201?}.
In all of these simulations, we simulated a population of 500 diploids for $10^4$ generations.
Each haploid genome consisted of a single genomic element $L=10^8$ basepairs long with recombination rate per basepair $r = 10^{-8}$ and overall mutation rate $\mu = 10^{-7}$.
We simulated two types of mutations: neutral mutations and beneficial mutations with additive effects and selection coefficient $s=0.1$.
With these parameters, $2Ns = 100$, so beneficial mutations are strongly selected and will sweep in $T_{\text{sweep}}\sim s^{-1} \log Ns \approx 50$ generations.
Such sweeps will effect a region of the chromosome $d_{\text{sweep}} \sim (r T_{\text{sweep}})^{-1} \approx 2 \times 10^6$ basepairs long.
Thus $d_{\text{sweep}} \ll L$, which will minimize edge effects of having a finite chromosome.

@@

The DPGP3 data set consists of haploid consensus sequences from $\sim 200$ files, obtained via the haploid embryo method of \cite{LangleyEtal2011}.
The SNP calls underlying these sequences have been subjected to a variety of quality filters described in \cite{LackEtAl2015}.
We obtained the DPGP3 consensus sequence files version 1.1 from \url{www.johnpool.net/genomes.html}.
These files contain the sequence alignments of all flies in the sample on all chromosome arms.
We also downloaded the Nov. 3, 2016 spreadsheet of inversions available at the link above.
For each chromosome arm, we excluded any samples with an inversion in that arm and then down-sampled to $n=100$ by taking the first 100 remaining samples in alphanumeric order by sample name.
(As a result, the data for each chromosome arm is from a slightly different subset of the individuals.)

@@

In addition to the average SFS used to fit the model, we computed the average 2-SFS for pairs of sites at distances distances between 3~bp and 5~Kb.
Because we are using four-fold degenerate sites, we only computed the 2-SFS for distances that are multiples of 3.
For comparison between data and simulations, we scale the distances in basepairs by a critical distance $d_c = (r\E{T_2})^{-1}$.
We used a genome-wide recombination rate of $r = 2 \times 10^{-8}$ per basepair per generation \autocite{ComeronEtAl2012}.
We estimated $\E{T_2}$ by $\Pi / 2 \mu$.
For $\Pi$, we used the average pairwise diversity at fourfold degenerate sites in the central region of each chromosome arm.
For $\mu$, we used a genome-wide mutation rate of $3 \times 10^{-9}$ per basepair per generation \autocite{KeightlyEtAl2014}.
These estimates are not precise, but only serve to scale genetic distances to the right order of magnitude.

@@

In this section, we compare the $\fpmi$ of the Kingman coalescent with and without population growth to the $\fpmi$ of the beta coalescent, for pairs of sites without recombination.
We are interested in whether $\fpmi$ can distinguish among coalescent models that produce similar distortions in the SFS.
To this end, we compute a version of Tajima's D \autocite{Tajima1983} normalized to be invariant under changes in the mean coalescent rate: $D = (\Pi - \hat{\theta}_W) / \Pi$, where $\hat{\theta}_W$ is Watterson's theta, a linear function of the SFS \autocite{Watterson19??}.
Negative values of $D$ indicate an overabundance of low-frequency mutations relative to the time-homogeneous Kingman expectation.
All results are computed numerically using the results of \cite{Fu1995}, \cite{ZivkovicWiehe2008}, and \cite{BirknerEtAl2013} (see Methods).

@@

The previous sections have shown that $\fpmi$ between pairs of non-recombining sites can discriminate between the Kingman coalescent with population growth and the beta coalescent.
However, most demographic inference is performed on regions of the genome with non-zero recombination rates.
In fact, for large non-recombining loci, it is sometimes possible to reconstruct the gene genealogy exactly and calculate tree imbalance statistics directly (see e.g., \cite{Seger2010}).
The advantage of the $\fpmi$ approach is that it can be computed on a genomic scale, as with the SFS.
Therefore, it is important to assess the robustness of our approach to recombination between sites.

@@

There is one other salient feature of \fig{fig:hilo_vs_d}: $\hilopmi$ does not decay to zero at long distances in the beta coalescent.
This is related to the results of \cite{EldonWakeley20??}, who showed that a model with ``jackpot'' reproductive events can generate infinite-range linkage disequilibrium.
They showed that different scalings of rates of mutation, pairwise coalescence, multiple merger coalescence, and recombination lead to different behaviors of diversity and linkage disequilibrium.
Our implementation of the beta coalescent model with recombination corresponds to a particular scaling limit where recombination does not have time to decorrelate trees during multiple mergers events, even at infinite genetic distances.
We do not expect this behavior to be universal to multiple mergers coalescents with recombination.

@@

Our results above show that $\fpmi$ and its binned analog, $\hilopmi$ are useful for distinguishing population growth from the effects of multiple mergers.
This is true even when the population growth generates similar distortions in the SFS.
We therefore propose the following model-checking procedure for demographic inference methods
\begin{enumerate}[label=(\roman*), before=\unskip{: }]
    \item Fit a demographic model to data.
    \item Simulate genealogies under the fitted model (using \texttt{msprime} or other coalescent simulator). Calculate the $\fpmi(d)$ and $\hilopmi(d;i_c)$ predicted by the model.
    \item Calculate $\fpmi(d)$ and $\hilopmi(d;i_c)$ from the data.
    \item Compare true to predicted statistics. Evaluate model fit.
\end{enumerate}
This procedure checks whether the model is consistent with a dimension the data that was used in fitting the model.
Inconsistency suggests that the inferred $N(t)$ may be an artifact of natural selection, skewed offspring distributions, etc., rather than reflecting the true historical population size.

@@

We have shown that $\fpmi$ and its binned analog $\hilopmi$ are sensitive to multiple mergers, but relatively invariant under population growth in the Kingman coalescent.
These properties make them well-suited for coalescent model checking.
We've demonstrated a model-checking procedure on data from \textit{D. melanogaster}, which is believed to be strongly shaped by natural selection, and found evidence that population growth can not explain the positive associations between high and low frequency mutations.

>>>>

One approach to identifying multiple mergers in genomic data is to use the SFS as a summary statistic.
\cite{BirknerEtAl2013, BlathEtAl2016, SpenceEtAl2016} derived methods for computing the expected site frequency spectrum of (simultaneous-)multiple merger coalescents.
\cite{EldonEtAl2015} showed that it is possible to distinguish between a multiple merger coalescent of the beta family and the Kingman coalescent with exponential growth using the SFS.
\cite{RodelspergerEtAl2014} detected widespread linked selection in the nematode \textit{Pristionchus pacificus} by demonstrating that the site frequency spectrum is non-monotonic, a signature of multiple mergers \autocite{NeherHallatscheck2013, BirknerEtAl2013}.

@@

Following the notation of \cite{Fu1995}, the site frequency spectrum of a sample of $n$ haploid genomes is $\left\{ \xi_i : 1 \leq i < n \right\}$, where $\xi_i$ is the fraction of sites containing a mutation with derived allele count $i$ in the sample.
In many cases, the ancestral allele is unknown and so the allele in $i$ samples and the complementary allele in $n-i$ samples are indistinguishable.
Therefore, we will mostly consider the \textit{folded} site frequency spectrum $\left\{ \eta_i = \xi_i + (1-\delta_{i,n-i}) \xi_{n-i}: 1 \leq i \leq \floor{n/2} \right\}$, where $\delta_{k,k'}$ is the Kronecker delta. %, where $\eta_i = \xi_i + \delta_{n-2i} \xi_{n-i}$.
The SFS and folded SFS can be calculated from a set of single nucleotide polymorphisms (SNPs) without any information about their relative locations in the genome.

@@

\cite{BirknerEtAl2013} extended Fu's calculation to a family of multiple merger coalescents called beta coalescents.
This one-parameter family interpolates between the Kingman coalescent and the Bolthausen-Sznitman coalescent as the parameter, $\alpha$, varies from 2 to 1.
Beta coalescents arise in models with fat-tailed offspring distributions \autocite{Schweinsberg2003}, and the Bolthausen-Sznitman coalescent is the limiting distribution of genealogies in rapidly adapting populations \autocite{NeherHallatscheck2013}.
Like Fu, Birkner et al.\ were primarily concerned with computing the sample variance of SFS-based summary statistics such as Tajima's D \autocite{Tajima1989}.
As a result, they were mostly interested in the diagonal terms of the SFS covariance matrix, which dominate that calculation.
However, Figures 5 and 6 of \cite{BirknerEtAl2013} show positive correlations between $\xi_i$ and $\xi_j$ for $j \not\in \{i, n-i\}$.
Thus, unlike the standard Kingman coalescent, the beta coalescent can generate positive associations between mutations with different minor allele counts.

@@

In our setting, the 2-SFS may be interpreted as a joint probability mass function over minor allele counts.
Given a coalescent model, the minor allele count at an arbitrary site is a random variable over $\{0,\ldots,\floor{n/2}\}$ with probability mass function $p(i) = \E{\eta_i}$, where we define $\eta_0$ to be the fraction of monomorphic sites.
Similarly, the minor allele counts at a pair of sites separated by $d$ bases is a pair of random variables with joint probability mass function $p_d(i,j) = \E{\eta_{ij}(d)}$.
Comparing, \eq{eq:fpmi} with \eq{eq:pmi} shows that $\fpmi$ is the standard pointwise mutual information for pairs of minor allele counts.

This transformation of the 2-SFS has several useful properties.
First, because it is based on the minor allele frequencies, we may compute $\fpmi$ from data without knowing the ancestral allele.
Moreover, the results of \cite{Fu1995} imply that $\fpmi(i,j) < 0$ for $i\neq j$ for non-recombining sites under the time-homogeneous Kingman coalescent.

@@

\subsection*{Weighted $\fpmi$}
For plotting purposes, we will also use a weighted version of the $\fpmi$:
\begin{align}
    \wfpmi(i,j:d) = \frac{\E{\eta_{ij}}}{\mu^2 \E{T_2}^2} \fpmi(i,j;d),
    \label{eq:wfpmi}
\end{align}
where $T_2$ is the coalescence time for a sample of size two.
The numerator of the weighting factor in \eq{eq:wfpmi} serves to emphasize the most common pairs of minor allele counts.
The denominator, which is proportional to the square of the expected pairwise diversity, $\Pi$, ensures that $\wfpmi$ is invariant under changes in the overall mutation and coalescence rates.

@@

For each coalescent model, we simulated at least $10^4$ independent infinite-sites loci, each with total recombination rate, $r$ (See SI for parameter combinations).
We chose values of $r$ so that the combined parameter $r\E{T_2}$ varied over several orders of magnitude, where $\E{T_2}$ was measured from the simulation output.
For each locus, we measured $\{\tau_i : i = 1,\ldots,n-1\}$ in the two genealogies at each end of the loci.
We then calculated the expectations $\{\E{\tau_i}\}$ and $\{\E{\tau_{ij}}\}$ by averaging over independent loci.
These expectations allow us to calculate all of the 2-SFS statistics defined above.

@@

We simulated a model of recurring selective sweeps using the software \texttt{SLiM} \autocite{MesserEtAl201?}.
In all of these simulations, we simulated a population of 500 diploids for $10^4$ generations.
Each haploid genome consisted of a single genomic element $L=10^8$ basepairs long with recombination rate per basepair $r = 10^{-8}$ and overall mutation rate $\mu = 10^{-7}$.
We simulated two types of mutations: neutral mutations and beneficial mutations with additive effects and selection coefficient $s=0.1$.
With these parameters, $2Ns = 100$, so beneficial mutations are strongly selected and will sweep in $T_{\text{sweep}}\sim s^{-1} \log Ns \approx 50$ generations.
Such sweeps will effect a region of the chromosome $d_{\text{sweep}} \sim (r T_{\text{sweep}})^{-1} \approx 2 \times 10^6$ basepairs long.
Thus $d_{\text{sweep}} \ll L$, which will minimize edge effects of simulating a finite chromosome.

@@

We analyzed sequence data from the DPGP3 data set, which consists of haploid consensus sequences from $\sim 200$ flies, obtained via the haploid embryo method of \cite{LangleyEtal2011}.
The SNP calls underlying these sequences have been subjected to a variety of quality filters described in \cite{LackEtAl2015}.
We obtained the DPGP3 consensus sequence files version 1.1 from \url{www.johnpool.net/genomes.html}.
These files contain the sequence alignments of all flies in the sample on all chromosome arms.
We also downloaded the Nov. 3, 2016 spreadsheet of inversions available at the link above.
For each chromosome arm, we excluded any samples with an inversion in that arm and then down-sampled to $n=100$ by taking the first 100 remaining samples in alphanumeric order by sample name.
(As a result, the data for each chromosome arm is from a slightly different subset of the individuals.)

@@

In addition to the average SFS used to fit the model, we computed the average 2-SFS for pairs of sites at distances distances between 3~bp and 5~Kb.
Because we are using four-fold degenerate sites, we only computed the 2-SFS for distances that are multiples of 3.
For comparison between data and simulations, we scale the distances in basepairs by a critical distance $d_c = (r\E{T_2})^{-1}$.
We used a genome-wide recombination rate of $r = 2 \times 10^{-8}$ per basepair per generation \autocite{ComeronEtAl2012}.
We estimated $\E{T_2}$ by $\Pi / 2 \mu$.
For $\Pi$, we used the average pairwise diversity at fourfold degenerate sites in the central region of each chromosome arm.
For $\mu$, we used a genome-wide mutation rate of $3 \times 10^{-9}$ per basepair per generation \autocite{KeightlyEtAl2014}.
These estimates are not precise, but only serve to scale genetic distances to the correct order of magnitude.

@@

We compared the $\fpmi$ of the Kingman coalescent with and without population growth to the $\fpmi$ of the beta coalescent, for pairs of sites without recombination.
We are interested in whether $\fpmi$ can distinguish among coalescent models that produce similar distortions in the SFS.
To this end, we computed a version of Tajima's D \autocite{Tajima1983} normalized to be invariant under changes in the mean coalescent rate: $D = (\Pi - \hat{\theta}_W) / \Pi$, where $\hat{\theta}_W$ is Watterson's theta, a linear function of the SFS \autocite{Watterson19??}.
Negative values of $D$ indicate an overabundance of low-frequency mutations relative to the time-homogeneous Kingman expectation.
All results are computed numerically using the results of \cite{Fu1995}, \cite{ZivkovicWiehe2008}, and \cite{BirknerEtAl2013} (see Methods).

@@

The previous sections have shown that $\fpmi$ between pairs of non-recombining sites can discriminate between the Kingman coalescent with population growth and the beta coalescent.
However, most demographic inference is performed on regions of the genome with non-zero recombination rates.
In fact, for large non-recombining loci, it is sometimes possible to reconstruct the gene genealogy explicitly and calculate tree imbalance statistics (see e.g., \cite{Seger2010}).
The advantage of the $\fpmi$ approach is that it can be computed on a genomic scale, as with the SFS.
Therefore, it is important to assess the robustness of our approach to recombination between sites.

@@

There is one other salient feature of \fig{fig:hilo_vs_d}: $\hilopmi$ does not decay to zero at long distances in the beta coalescent.
This is related to the results of \cite{EldonWakeley20??}, who showed that a model with ``jackpot'' reproductive events can generate infinite-range linkage disequilibrium.
They showed that different scalings of rates of mutation, pairwise coalescence, multiple merger coalescence, and recombination lead to different behaviors of diversity and linkage disequilibrium.
Our implementation of the beta coalescent model with recombination corresponds to a particular scaling limit where recombination does not have time to decorrelate trees during multiple mergers events, even at infinite genetic distances.
We do not expect this behavior to be universal in multiple mergers coalescents with recombination.

@@

Our results above show that $\fpmi$ and its binned analog, $\hilopmi$ are useful for distinguishing population growth from the effects of multiple mergers.
This is true even when the population growth generates similar distortions in the SFS.
We therefore propose the following model-checking procedure for demographic inference methods
\begin{enumerate}[label=(\roman*), before=\unskip{: }]
    \item Fit a demographic model to data.
    \item Simulate genealogies under the fitted model (using \texttt{msprime} or other coalescent simulator). Calculate the $\fpmi(d)$ and $\hilopmi(d;i_c)$ predicted by the model.
    \item Calculate $\fpmi(d)$ and $\hilopmi(d;i_c)$ from the data.
    \item Compare true to predicted statistics. Evaluate model fit.
\end{enumerate}
This procedure checks whether the model is consistent with a dimension the data that was not used in fitting the model.
Inconsistency suggests that the inferred $N(t)$ may be an artifact of natural selection, skewed offspring distributions, etc., rather than reflecting the true historical population size.

@@

We have shown that $\fpmi$ and its binned analog $\hilopmi$ are sensitive to multiple mergers, but relatively invariant under population growth in the Kingman coalescent.
These properties make them well-suited for coalescent model checking.
We have demonstrated a model-checking procedure on data from \textit{D. melanogaster}, which is believed to be strongly shaped by natural selection, and found evidence that population growth can not explain the positive associations between high and low frequency mutations.
